I Compensate, therefore I Am \\ (accurate for DNN verification)

### 2024-01-20 Paper/Paper_noProof.tex
bibliography

\begin{table}[b!]
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
		\text{Source/Target Layers}  &  \text{Natural DNN} & \text{Robust DNN} & \text{Ratio Natural vs Robust} \\ \hline \hline
	0 / 2 & 0.0304 & 0.00220  & 13.8x\\ \hline
	1 / 3  & 0.0313 & 0.00875 & 3.58x \\ \hline
	2 / 4  &  0.0267 & 0.00785 & 3.40x \\ \hline
	3 / 5  &  0.0253 & 0.00804  & 3.18x \\ \hline
\end{tabular}
\caption{Comparison of the average compensation strength over all the pairs of nodes of layer source/target between a Network Naturally trained and a Network Robustly trained with the same architecture and the same training set.}
\label{tab:compensation}
\end{table}

@@

This result has a first interesting application: explain why some DNNs are intrinsically easier to verify than others. This is the case for DNNs trained adversarially to be robust, vs same architecture, but trained in a "natural" way \cite{deeppoly,prima,crown}. Indeed, compensating pairs is a purely structural property on the DNN learnt, rather than more semantical notions, such as the number of unstable ReLUs, which depends heavily upon the image considered and the algorithm used. We compare compensations with source and target separated by 1 ReLU layer, e.g. source in layer 0 and target in layer 2, for two Networks with 5 hidden layers of 100 nodes each from the ERAN GitHub repository, one trained naturally ("$6\times100$"), 
and one trained using DiffAI ("$5\times100$"). 
We report in Table \ref{tab:compensation} the average compensation strength, i.e. 
the average of $max_{\pi,\pi'} min(|weight(\pi')|,|weight(\pi')|)$
over all source/target states in given layers. The average compensations strength are much lower in the Robust DNN than in the Natural DNN, and indeed the Robust DNN is much simpler to verify than the Natural DNN (e.g. DeepPoly is more accurate and verifies more images).

@@

The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes, to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.

@@

In terms of efficient verifiers we compare with, we report:
\begin{itemize}
	\item DeepPoly \cite{deeppoly}/ CROWN \cite{crown}. We report the runtime from our own implementation used in Compensate($K$). It shows that our purely Python implementation is not optimal ($10$-$100$ times slower than the implementation in ($\alpha$)($\beta$)CROWN, ERAN and PRIMA). Our aim is indeed not to have the most optimized tool, but to evaluate the accuracy of Compensation.
	\item PRIMA, with results reported in \cite{prima} and \cite{crown}, as it is the closest in practice with our algorithm, computing accurate bounds for all neurons (as we do), although with a different technique: they focus on explicit correlations between neurons, while we consider the correlations indirectly through a better network abstraction. They use GPU while we do not. When computing the whole correlated space, 
	parallelization is difficult.
	Notice that on these MLP networks, PRIMA resort to refined bounds on the first few layers (3 or 4) using exact MILP encoding of all the nodes (which is easily parallelized - 16 nodes considered in parallel in PRIMA). This is doable on the first few layers as the number of integer variable is not too large. By comparison, we run MILP on all the layers, but with a bounded number of integer variables to control the runtime and to scale to all the layers.
	\item ($\alpha$)$\beta$-CROWN, with results as reported in \cite{crown}, as it is the most accurate of the efficient tool to date. Its main idea is to implement a very efficient Branch and Bound (BaB) verification tool, subsuming \cite{BaB}, and heavily using GPU implementation (while we do not). Performance of pure BaB on these hard DNNs is however impaired by the large number of branchs to consider. Similarly as PRIMA, $\beta$-CROWN resorts to refining the bounds of the first few layers using an exact MILP encoding, also computing 16 nodes in parallel, before running their BaB tool. 
	\item We also experimented with the latest versions of $\alpha$-$\beta$-CROWN (October 2023) and PRIMA (2022) on MLP $6 \times 500$ (results not reported in \cite{crown,prima}).
	\item We do not report results from kPoly \cite{kpoly}, OptC2V \cite{optC2V}, 
	SDPFO \cite{SDPFI} which do not present favorable trade offs compared with $\beta$-CROWN. Results can be found in \cite{crown} and are performed in the same conditions so can be compared.
	\item Finally, we run several version of Compensate($K$) ({\color{blue} Comp} in Table \ref{tab:example}). First, we have a fixed choice for $K$ ($K=60$ for fast and $K=200$ for slow), with a fall back to a lower $K$ ($K=50$ for fast and $K=60$ for slow) if the bounds found is too far from the solution found (MIPGap $>0.5$ in Gurobi). We run (our implementation of) DeepPoly first, then the fast version, before running slow on the remaining images.
	Further, we also experiment with an {\em adaptative} implementation, which lowers number $K$ adaptively between layers to adapt to the growing number $N$ of linear variables, which has an impact on the overall $\MILP(N,K)$ runtime. This is particularly useful for deeper network where $N$ is very different between the first and the last layer. Again, we run first an aggressive version of fast, before going to a more conservative slow version to treat the remaining images. All versions compute bounds for 20 neurons of the same layer in parallel, no GPU used.
\end{itemize}

>>>>

\begin{table}[b!]
	\centering
	\begin{tabular}{|c|c|c|c|}
	\hline
		\text{Source/Target Layers}  &  \text{Natural DNN} & \text{Robust DNN} & \text{Ratio Natural vs Robust} \\ \hline \hline
	0 / 2 & 0.0304 & 0.00220  & 13.8x\\ \hline
	1 / 3  & 0.0313 & 0.00875 & 3.58x \\ \hline
	2 / 4  &  0.0267 & 0.00785 & 3.40x \\ \hline
	3 / 5  &  0.0253 & 0.00804  & 3.18x \\ \hline
\end{tabular}
\caption{Comparison of the average compensation strength over all the pairs of nodes of layer source/target between a DNN naturally-trained and a DNN robustly-trained using DiffAI \cite{DiffAI} with the same architecture and training set.}
\label{tab:compensation}
\end{table}

@@

This result has a first interesting application: explain why some DNNs are intrinsically easier to verify than others. This is the case for DNNs trained adversarially to be robust, vs same architecture, but trained in a "natural" way \cite{deeppoly,prima,crown}. Indeed, compensating pairs is a purely structural property on the DNN learnt, rather than more semantical notions, such as the number of unstable ReLUs, which depends heavily upon the image considered and the algorithm used. We compare compensations with source and target separated by 1 ReLU layer, e.g. source in layer 0 and target in layer 2, for two Networks with 5 hidden layers of 100 nodes each from the ERAN GitHub repository, one trained naturally ("$6\times100$"), 
and one trained using DiffAI \cite{DiffAI} ("$5\times100$"). 
We report in Table \ref{tab:compensation} the average compensation strength, i.e. 
the average of $max_{\pi,\pi'} min(|weight(\pi')|,|weight(\pi')|)$
over all source/target states in given layers. The average compensations strength are much lower in the Robust DNN than in the natural DNN, and indeed the Robust DNN is much simpler to verify than the natural DNN (e.g. DeepPoly is more accurate and verifies more images).

@@

The first group of experiments we run, tests the accuracy of choosing a set $Z$ of $K$ nodes using compensation strength, compared with randomly choosing the same number $K$ of nodes (similar to MIPplanet \cite{MIPplanet}),  to understand if the choice is meaningful or any set $Z$ with the same size will result in a similar accuracy.

@@

In terms of efficient verifiers we compare with, we report:
\begin{itemize}
	\item DeepPoly \cite{deeppoly}/ CROWN \cite{crown}. We report the runtime from our own implementation used in Compensate($K$). It shows that our purely Python implementation is not optimal ($10$-$100$ times slower than the implementation in ($\alpha$)($\beta$)CROWN, ERAN and PRIMA). Our aim is indeed not to have the most optimized tool, but to evaluate the accuracy of Compensation.
	\item PRIMA, with results reported in \cite{prima} and \cite{crown}, as it is the closest in practice with our algorithm, computing accurate bounds for all neurons (as we do), although with a different technique: they focus on explicit correlations between neurons, while we consider the correlations indirectly through a better network abstraction. They use GPU while we do not. When computing the whole correlated space, 
	parallelization is difficult.
	Notice that on these MLP networks, PRIMA resort to refined bounds on the first few layers (3 or 4) using exact MILP encoding of all the nodes (which is easily parallelized - 16 nodes considered in parallel in PRIMA). This is doable on the first few layers as the number of integer variable is not too large. By comparison, we run MILP on all the layers, but with a bounded number of integer variables to control the runtime and to scale to all the layers.
	\item ($\alpha$)$\beta$-CROWN, with results as reported in \cite{crown}, as it is the most accurate of the efficient tool to date. Its main idea is to implement a very efficient Branch and Bound (BaB) verification tool, subsuming \cite{BaB}, and heavily using GPU implementation (while we do not). Performance of pure BaB on these hard DNNs is however impaired by the large number of branchs to consider. Similarly as PRIMA, $\beta$-CROWN resorts to refining the bounds of the first few layers using an exact MILP encoding, also computing 16 nodes in parallel, before running their BaB tool. 
	\item We also experimented with the latest versions of $\alpha$-$\beta$-CROWN (October 2023) and PRIMA (2022) on MLP $6 \times 500$ (results not reported in \cite{crown,prima}).
	\item We do not report results from kPoly \cite{kpoly}, OptC2V \cite{optC2V}, 
	SDPFO \cite{SDPFI}, MIPplanet \cite{MIPplanet} which do not present favorable trade offs compared with $\beta$-CROWN. Results can be found in \cite{crown} and are performed in the same conditions so can be compared.
	\item Finally, we run several version of Compensate($K$) ({\color{blue} Comp} in Table \ref{tab:example}). First, we have a fixed choice for $K$ ($K=60$ for fast and $K=200$ for slow), with a fall back to a lower $K$ ($K=50$ for fast and $K=60$ for slow) if the bounds found is too far from the solution found (MIPGap $>0.5$ in Gurobi). We run (our implementation of) DeepPoly first, then the fast version, before running slow on the remaining images.
	Further, we also experiment with an {\em adaptative} implementation, which lowers number $K$ adaptively between layers to adapt to the growing number $N$ of linear variables, which has an impact on the overall $\MILP(N,K)$ runtime. This is particularly useful for deeper network where $N$ is very different between the first and the last layer. Again, we run first an aggressive version of fast, before going to a more conservative slow version to treat the remaining images. All versions compute bounds for 20 neurons of the same layer in parallel, no GPU used.
\end{itemize}

### 2024-01-20 Paper/introduction.tex
bibliography

In this context, application of DNNs in safety critical applications is cautiously envisioned. For that to happen at a large scale, hard guarantees should be provided, so that to avoid dramatic consequences. It is the reason for the development of (hard) verification tools since 2016, with now many tools with different trade-offs from exact computation but slow (e.g. Marabou \cite{katz2019marabou}/Reluplex\cite{Reluplex}), up to very efficient but also incomplete (e.g. ERAN-DeepPoly \cite{deeppoly}). To benchmark these tools, a competition has been run since 2019, namely VNNcomp. The current overall better performing verifier is $\alpha$-$\beta$-Crown \cite{crown}, a fairly sophisticatedly engineered tool based mainly on "branch and bound" (BaB), and which can scale all the way from complete on smaller DNNs \cite{xu2020fast} up to very efficient on larger DNNs, constantly upgraded, e.g. \cite{cutting}.

While the verification engines are generic, the benchmarks usually focus on local robustness, i.e. given a Network, an image and a small neighbourhood around this image, 
is it the case that all the images in the neighbourhood are classified in the same way.
While some quite large DNNs (e.g. ResNet with tens of thousands of neurons) can be verified very efficiently (tens of seconds per input) \cite{crown}, with all inputs either certified robust or an attack on robustness is found; some smaller DNNs (with hundreds of neurons, only using the simpler ReLU activation function) cannot be analysed fully, with $12-20\%$ of inputs where neither of the decisions can be reached \cite{crown} and Table \ref{tab:example}. The main difference between these DNNs with very different behaviours lies in that the DNNs which are trained to be robust are much easier to verify, while the DNNs trained in a "natural" way are much harder to verify.

In this paper, we focus on DNNs trained in a "natural", 
%uncovering what makes the DNNs trained in a natural way so hard to verify (
because for "easier" DNNs, adequate methods already exist. 
To do so, we analyse the abstraction mechanisms at the heart of several efficient algorithms, namely Eran-DeepPoly \cite{deeppoly}, the Linear Programming approximation \cite{MILP}, PRIMA \cite{prima}, and different versions of ($\alpha$)($\beta$)-CROWN \cite{crown}. All these algorithms compute lower or/and upper bounds for the values of neurons (abstraction on values) for inputs in the considered input region, and conclude based on such bounds. For instance, if for all image $I'$ in the neighbourhood of image $I$, we have $weight_{I'}(n'-n) < 0$ for $n$ the output neuron corresponding to the expected class, then we know that the DNN is robust in the neighbourhood of image $I$. We restrict the formal study to DNNs using only the standard ReLU activation function, although nothing specific prevents the results to be extended to more general architectures. We uncover that {\em compensations} 
(see next paragraph) is the phenomenon creating inaccuracies. We verified experimentally that DNNs trained in a natural way have much more heavy compensating pairs than DNNs trained in a robust way.

@@

For comparison, $\alpha$-$\beta$ CROWN \cite{crown} (and other Branch and Bound algorithms, such as BaB \cite{BaB}) will run few instances of branch and bound (one per output neuron), in worst case considering all the possible ReLU configurations (although the branch and bound algorithm avoids most of the possibilities). On simple networks, such as those trained robust, this is particularly efficient because branch and bound can find very efficiently the bounds focusing on the actual question, considering the important branches, while our algorithm will be less efficient as it has to consider each node one by one from the start. However, branch and bound faces a complexity wall when the network is hard to verify, such as the DNNs trained naturally, as there are too many branches to consider.

>>>>

In this context, application of DNNs in safety critical applications is cautiously envisioned. For that to happen at a large scale, hard guarantees should be provided \cite{certification}, through e.g. incremental verification \cite{incremental}, so that to avoid dramatic consequences. It is the reason for the development of (hard) verification tools since 2016, with now many tools with different trade-offs from exact computation but slow (e.g. Marabou \cite{katz2019marabou}/Reluplex\cite{Reluplex}), up to very efficient but also incomplete (e.g. ERAN-DeepPoly \cite{deeppoly}). To benchmark these tools, a competition has been run since 2019, namely VNNcomp \cite{VNNcomp}. The current overall better performing verifier is $\alpha$-$\beta$-Crown \cite{crown}, a fairly sophisticatedly engineered tool based mainly on "branch and bound" (BaB) \cite{BaB}, and which can scale all the way from complete on smaller DNNs \cite{xu2020fast} up to very efficient on larger DNNs, constantly upgraded, e.g. \cite{cutting}.

While the verification engines are generic, the benchmarks usually focus on local robustness, i.e. given a DNN, an image and a small neighbourhood around this image, 
is it the case that all the images in the neighbourhood are classified in the same way.
While some quite large DNNs (e.g. ResNet with tens of thousands of neurons) can be verified very efficiently (tens of seconds per input) \cite{crown}, with all inputs either certified robust or an attack on robustness is found; some smaller DNNs (with hundreds of neurons, only using the simpler ReLU activation function) cannot be analysed fully, with $12-20\%$ of inputs where neither of the decisions can be reached (\cite{crown} and Table \ref{tab:example}). Actually, DNNs which are trained to be robust (using DiffAI \cite{DiffAI} or PGD \cite{PGD}) are easier to verify, while the DNNs trained in a "natural" way are harder to verify.

In this paper, we focus on DNNs trained in a "natural" way,
%uncovering what makes the DNNs trained in a natural way so hard to verify (
because for "easier" DNNs, adequate methods already exist. 
To do so, we analyse the abstraction mechanisms at the heart of several efficient algorithms, namely Eran-DeepPoly \cite{deeppoly}, the Linear Programming approximation \cite{MILP}, PRIMA \cite{prima}, and different versions of ($\alpha$)($\beta$)-CROWN \cite{crown}. All these algorithms compute lower or/and upper bounds for the values of neurons (abstraction on values) for inputs in the considered input region, and conclude based on such bounds. For instance, if for all image $I'$ in the neighbourhood of image $I$, we have $weight_{I'}(n'-n) < 0$ for $n$ the output neuron corresponding to the expected class, then we know that the DNN is robust in the neighbourhood of image $I$. We restrict the formal study to DNNs using only the standard ReLU activation function, although nothing specific prevents the results to be extended to more general architectures. We uncover that {\em compensations} 
(see next paragraph) is the phenomenon creating inaccuracies. We verified experimentally that a DNN trained in a natural way has heavier compensating pairs than DNNs trained in a robust way.

@@

For comparison, $\alpha$-$\beta$ CROWN \cite{crown} (and other Branch and Bound algorithms, such as BaBSR \cite{BaB}) will run few instances of branch and bound (one per output neuron), in worst case considering all the possible ReLU configurations (although the branch and bound algorithm avoids most of the possibilities). On simple networks, such as those trained robust, this is particularly efficient because branch and bound can find very efficiently the bounds focusing on the actual question, considering the important branches, while our algorithm will be less efficient as it has to consider each node one by one from the start. However, branch and bound faces a complexity wall when the network is hard to verify, such as the DNNs trained naturally, as there are too many branches to consider.
